{\scshape Denicek}: Programming Substrate for Concrete, \\Collaborative, Interactive Programming

### 2024-11-01 paper.tex
Explain primitive operations, first draft

@@

@@

\begin{tabular}{|clp{18.08em}|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
\sffamily\small{\bfseries Selector} & {\sffamily\footnotesize Notation} & \\[0.2em]
\hline
&&\\[-1em]
\seltablecol{\Verb|field|}{Field}{Refers to record field of a given name}
\seltablecol{\Verb|number|}{Index}{Refers to list element at a given index}
\seltablecol{\Verb|*|}{Any}{Refers to all children of a list node}
\hline
\end{tabular}

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Node}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, child$_1$, $\ldots$, child$_n$}
  {Ordered list of nodes, addressable by index. Renders as \Verb|<tag>| with children}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field} name. Renders as \Verb|<tag>|.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \Verb|/selectors| as a link.}
\ndtablecol{\faFont}{Primitivie}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Structure of selectors and document nodes.}
\label{fig:doc}
\end{figure}

@@

\section{The Denicek Substrate}
\label{sec:system}
Denicek represents programs as a sequence of edits construct a computational document.
In this section, we describe the structure of documents and edits, as well as the
operations that form the backbone of the system and are used to implement a range of
interesting programming experiences, discussed in \S\ref{sec:impl}.

\subsection{Selectors, Documents and Edits}
Denicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
computational document visible to the user, although parts can be collapsible or hidden using CSS.
Records and lists are rendered as HTML elements of a specified tag with children becoming child
elements. Field names are hidden in the rendered document.

References to document location are used in both document itself (reference nodes) and in edits
(target and a source). They are represented as a sequence of selectors (Fig.~\ref{fig:doc}).
The document model assumes that lists are homogeneous and records heterogeneous. Thus the
\textit{Any} selector makes it possible to refer to all children of a list, but there is no
way to refer to all children of a record.

\subsubsection*{Document Edits}
The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
require \emph{target} to which they are applied. The target is specified through selectors.
Most edits can only be applied when the target is a node of a particular kind.

The edit operations (Fig.~\ref{fig:edits}) are designed to allow any transformation of a document
through a series of steps whose effect can be tracked by the substrate. As illustrated earlier,
Denicek updates selectors when document structure changes. Fig.~\ref{fig:edits} distinguishes
between edits that keep existing references in a document unchanged (above) and edits that affect
references (below). When a selector is \emph{invalidated}, e.g., when deleting a field or a
list item to which there is a reference, Denicek prevents the edit. Copying also invalidates
selectors because it is ambiguous whether selectors referring to the original location should
refer to the source or the target of the copying after the edit (and since reference can only
refer to one location, duplicating it is not possible). Updating a field, wrapping or reordering
requires changing existing references in document correspondingly.

@@

\begin{figure}
\newcommand{\ektablecol}[6]{
\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 & \sffamily\footnotesize #5 \\[-0.2em]
&\multicolumn{3}{l|}{\sffamily\footnotesize #6}\\[0.3em]
}
\begin{tabular}{|c|p{13em}rr|}
\hline
\rowcolor{ekgray}
&&&\\[-1em]
\rowcolor{ekgray}
 & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target & \sffamily\footnotesize\bfseries Selectors \\[0.2em]
\hline
&&&\\[-1em]
\ektablecol{\faPlus}{Add}{field, node}{Record}{Unchanged}
  {Add \textit{node} as a \textit{field} to the specified record.}
\ektablecol{\faAt}{Append}{node}{List}{Unchanged}
  {Append \textit{node} to the end of the specified list.}
\ektablecol{\faPaperclip}{AppendFrom}{selectors}{List}{Unchanged}
  {Append node from \emph{selectors} to the end of the specified list.}
\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}{Unchanged}
  {Apply primitive \textit{transform} to the specified primitive.}
\ektablecol{\faCode}{UpdateTag}{old tag, new tag}{Tagged}{Unchanged}
  {Change the tag of a specified list or record from \textit{old} to \textit{new}.}
\hline
&&&\\[-1em]
\ektablecol{\faFont}{UpdateField}{old field, new field}{Record}{Update}
  {Rename the field of a specified record from \textit{old} to \textit{new}.}
\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}{Invalidate}
  {Delete the field \textit{field} of a specified record.}
\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}{Invalidate}
  {Delete the item at a given \textit{index} of a specified list.}
\ektablecol{\faSort}{Reorder}{permutation}{List}{Update}
  {Reorder items of a specified list according to a \textit{permutation}.}
\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}{Update}
  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
\ektablecol{\faListUl}{WrapList}{tag}{Any}{Update}
  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
\ektablecol{\faCopy}{Copy}{selectors}{Any}{Invalidate}
  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Types of document edits.}
\label{fig:edits}
\end{figure}

@@

\subsubsection*{Automatic Reference Update}
There are three situations in which automatic update of references is undesirable. First, if
an edit is not applied to all elements of a list (reference contains the \emph{index} selector),
it should not affect references that refer into the list. Such edits may turn document into an
inconsistent state, but they typically do so temporarily during document construction
(checking such edits is discussed in \S\ref{sec:discuss-future}).

Second, documents can contain values that can be of multiple different kinds (i.e., a union type).
In such case, references should not be updated when the kind of the value changes. For example,
a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
involves wrapping, but this should not affect references to the formula.
To support those cases, it is possible to specify that edits that normally affect selectors
(Fig.~\ref{fig:edits}, below) should not do so. Moreover, this setting is required when the
target reference contains the \emph{index} selector.

\subsubsection*{Non-Conditionality of Edits.}
NON-CONDITIONALITY effect does not depend on document state

\subsection{Merging histories}
is tricky in various ways

\subsubsection*{Updating structure for structural edits}
yay

\subsection{Conflict detection}
is pretty easy

\subsection{Properties}

@@

3) Interaction (add button)

@@

>>>>

JUSTIFICATION - this is a technical paper but belongs here!

@@

The coolest trick - almost everything (PBD, collaboration, evaluation) is done through merge!

@@

\begin{tabular}{|llp{18.08em}|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
\sffamily\small{\bfseries Selector} & {\sffamily\footnotesize Notation} & \\[0.2em]
\hline
&&\\[-1em]
\seltablecol{\Verb|field|}{Field}{Refers to record field of a given name}
\seltablecol{\Verb|number|}{Index}{Refers to list element at a given index}
\seltablecol{\Verb|*|}{Any}{Refers to all children of a list node}
\hline
\end{tabular}

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, child$_1$, $\ldots$, child$_n$}
  {Ordered list of nodes, addressable by index. Renders as \Verb|<tag>| with children}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field} name. Renders as \Verb|<tag>|.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \Verb|/selectors| as a link.}
\ndtablecol{\faFont}{Primitivie}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\vspace{-1em}
\end{figure}

@@

\section{The Denicek Substrate}
\label{sec:system}
Denicek represents programs as sequences of edits that construct and transform a computational
document. In this section, we describe the structure of documents and edits, as well as the
operations that form the backbone of the system and are used to implement a wide range of
programming experiences, discussed in \S\ref{sec:impl}.

\subsection{Selectors, Documents and Edits}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
Denicek follows the \emph{naive realism}~\cite{disessa-1986-boxer} principle and makes the entire
document visible to the user, although parts can be collapsible or hidden using CSS.
Records and lists are rendered as HTML elements of a specified tag with children becoming child
elements. Field names are hidden in the rendered document.

References to document location are used in both document itself (reference nodes) and in edits
(target of the edit). They are represented as a sequence of selectors (Fig.~\ref{fig:doc}).
The document model assumes that lists are homogeneous and records heterogeneous, and so the
\textit{Any} selector makes it possible to refer to all children of a list, but there is no
way to refer to all children of a record.

\subsubsection*{Document Edits}
The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
require \emph{target} to which they are applied. Target is a reference and can contain the
\textit{Any} selector, in which case the edit is applied to multiple nodes simultaneously.
Most edits can only be applied to target node(s) of a particular kind.

The edits are designed to allow any transformation of a document
through a series of steps whose effect can be tracked by the substrate. As illustrated earlier,
Denicek updates references when document structure changes. Fig.~\ref{fig:edits} distinguishes
between edits that keep existing references in a document unchanged (above) and edits that affect
references (below). When a selector is \emph{invalidated}, e.g., when deleting a field or a
list item to which there is a reference, Denicek rejects the edit. Copying also invalidates
selectors because it is ambiguous whether selectors referring to the original location should
refer to the source or the target of the copying after the edit (and references cannot refer to
multiple structurally incompatible locations). Updating a field, wrapping, reordering or deleting
a list item requires updating references in the document correspondingly.

@@

\begin{figure}
\newcommand{\ektablecol}[6]{
\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 & \sffamily\footnotesize #5 \\[-0.2em]
&\multicolumn{3}{l|}{\sffamily\footnotesize #6}\\[0.3em]
}
\begin{tabular}{|cp{13em}rr|}
\hline
\rowcolor{ekgray}
&&&\\[-1em]
\rowcolor{ekgray}
 & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target & \sffamily\footnotesize\bfseries Selectors \\[0.2em]
\hline
&&&\\[-1em]
\ektablecol{\faPlus}{Add}{field, node}{Record}{Unchanged}
  {Add \textit{node} as a \textit{field} to the specified record.}
\ektablecol{\faAt}{Append}{node}{List}{Unchanged}
  {Append \textit{node} to the end of the specified list.}
\ektablecol{\faPaperclip}{AppendFrom}{selectors}{List}{Unchanged}
  {Append node from \emph{selectors} to the end of the specified list.}
\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}{Unchanged}
  {Apply primitive \textit{transform} to the specified primitive.}
\ektablecol{\faCode}{UpdateTag}{old tag, new tag}{Tagged}{Unchanged}
  {Change the tag of a specified list or record from \textit{old} to \textit{new}.}
\hline
&&&\\[-1em]
\ektablecol{\faFont}{UpdateField}{old field, new field}{Record}{Update}
  {Rename the field of a specified record from \textit{old} to \textit{new}.}
\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}{Invalidate}
  {Delete the field \textit{field} of a specified record.}
\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}{Update}
  {Delete the item at a given \textit{index} of a specified list.}
\ektablecol{\faSort}{Reorder}{permutation}{List}{Update}
  {Reorder items of a specified list according to a \textit{permutation}.}
\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}{Update}
  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
\ektablecol{\faListUl}{WrapList}{tag}{Any}{Update}
  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
\ektablecol{\faCopy}{Copy}{selectors}{Any}{Invalidate}
  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Summary of document edit types in Denicek}
\label{fig:edits}
\vspace{-1em}
\end{figure}

@@

\subsubsection*{Automatic Reference Update}
There are two situations in which automatic update of references is undesirable. If an edit is
applied to a singular element of a list (reference contains the \textit{Index} selector),
references that refer into any element of the list should be unchanged.  Such edits may turn
document into an inconsistent state, but they typically do so temporarily during document
construction (checking such edits is discussed in \S\ref{sec:discuss-future}).

Documents can also contain values that can be of multiple different kinds (i.e., a union type).
In such case, references should not be updated when the kind of the value changes. For example,
a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
involves wrapping, but this should not affect references to the formula.
To support those cases, it is possible to annotate edits that normally affect selectors
(Fig.~\ref{fig:edits}, below) as non-structural. This annotation is required when the
target reference contains the \emph{Index} selector.

\subsubsection*{No Conditional Edits.}
An important aspect of the design is that the effect an edit has on references inside
the document does not depend on the current value of the document. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
(The effect of \textit{Copy} depends only on the existing document structure, but not on its value;
while \textit{AppendFrom} affects only single list element and is thus labelled as a
non-structural edit.)

This design choice makes it impossible to encode computational logic directly in the edits
(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
has to be provided as an additional  mechanism on top of the underlying Denicek substrate.

\subsection{Primitive Operations}
Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
sequences of edits can be checked for conflicts or merged. Denicek identifies edit histories by
a (git-like) hash, computed from the hash of the parent and the current edit, which is used to identify
common shared part of the history during conflict checking and merging.

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transform it according to the
edit. If the edit may affect references in the document (Fig~\ref{fig:edits}, below), Denicek
updates matching references in the document according to the rules shown in Fig.~\ref{fig:updates}
provided that (i) the edit is not explicitly marked as non-structural, and (ii) the target
reference does not contain \textit{Index} selector, i.e., it targets all elements of any involved
lists. Also note that matching references in the document may be more specific than the edit
target. For example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|},
a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. If the document
contains a reference that would be invalidated by an edit, the edit is rejected.

\subsubsection*{Conflict Detection}
Two edits are conflicting if the order in which they are applied matters. This is the case if
they target the same node, or if one targets a node that is nested inside the node targetted by
the other. For edits with dependencies (\textit{AppendFrom} and \textit{Copy}), a conflict also
occurs if the other edit modifies the dependency.

Given two edit histories, one way to resolve conflicts is by removing (or marking as disabled) all
edits from one of the histories that conflict with edits done by the other history. To do this,
we first collect all targets of edits in the other history. We then iterate over edits from the first
history to see if they depend on or target any of the affected targets. If an edit is removed,
its target reference is added to the set of affected targets, so that other subsequent edits that depend
on its original result are also removed.

\subsubsection*{Merging Edit Histories}
Removing all conflicting edits is sometimes useful, but there are many cases where conflicting
edits, in the above sense, can be reconcilliated. Say we have edit histories with
the same common shared part $E, E_1$ and $E, E_2$. If the edits in $E_1$ and $E_2$ conflict,
the results of applying $E, E_1, E_2$ and $E, E_2, E_1$ would be different. We can, however,
construct edit histories $E_1'$ and $E_2'$ such that the results of applying $E, E_1, E_2'$ and
$E, E_2, E_1'$ are the same.

The key operation that enables such reconcilliation takes two individual edits that occurred
indpendently, $e_1$ and $e_2$, and produces $e_1'$ that has the same logical effect as
$e_1$, but can be applied after $e_2$. There are two aspects of such reconcilliation:

\begin{enumerate}
\item \emph{Apply Edit to Newly Added.} If $e_1$ is adding a new list item, but~$e_2$ is
  changing elements of the target list, we apply the edit $e_2$ to the node added by $e_1$ so
  that, when it is added after the transformation, the added node has the new structure.

\item \emph{Transform Matching References.} If $e_1$ targets a node that is inside a node
  whose structure is changed by $e_2$, the target reference in $e_1$ is updated in a way that
  corresponds to the new structure. That is, using the same rules, shown in Fig.~\ref{fig:updates},
  that apply when transforming references inside a document.
\end{enumerate}

Applying an edit to a newly added node involves a number of cases. When the new node is added
using \textit{Add} or \textit{Copy}, but the target location is modified, this is an unresolvable
conflict (because those two operations would overwrite existing nodes). When the new node is
added as a new list item using \textit{Append}, it can be transformed (by applying the other
edit direclty to the new node).

Finally, when the new node is added using \textit{AppendFrom},
it is copied from another document location. We cannot transform the node in the soource location
(this would have unintended effects) or after adding it (we do not know its new index because of
aforementioned \emph{non-conditionality} of edits). Denicek reconcilliates such edits by first
copying the source node to a new temporary location, transforming it there and then using
\textit{AppendFrom} from the temporary location.

\begin{figure}
\newcommand{\tttablecol}[5]{
\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
}
\begin{tabular}{|l|}
\hline
\\[-1em]
\tttablecol{UpdateField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
  {Replace Field for matching references.}
\tttablecol{DeleteItem}{target, index}{\Verb|/target/n/nested|}{\Verb|/target/(n-1)/nested|}
  {Decrement Index greater than $n$ in matching refs.}
\tttablecol{Reorder}{target, permutation}{\Verb|/target/n/nested|}{\Verb|/target/permutation(n)/nested|}
  {Update Index using permutation in matching refs.}
\tttablecol{WrapRecord}{target, tag, field}{\Verb|/target/nested|}{\Verb|/target/field/nested|}
  {Insert extra Field selector after matching prefix.}
\tttablecol{WrapList}{target, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
  {Insert extra All selector after matching prefix.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{How document edits transform references}
\label{fig:updates}
\vspace{-1em}
\end{figure}

@@

\subsection{Interaction (add button + remove done)}
\label{sec:impl-interaction}

@@

AppendFrom - hard to avoid! because recorded edits cannot know length
(but having -1 in index would break too)
